Concrete Semantics for Pushdown Analysis:\\The Essence of Summarization

### 2013-05-09 paper.tex
Shift-reset semantics in paper done. Added memo tables to cfa2 and pdcfa semantics.

\begin{align*}
  \mexpr \in \Expr &= \svar{\mvar} \mid \sapp{\mexpr}{\mexpr} \mid \slam{\mvar}{\mexpr} \\
  \mstate \in \State &= (\Expr \times \Env) \times \Store \times \Kont \\
  \mval \in \Value &::= \sclo{\slam{\mvar}{\mexpr}}{\menv} \\
  \mkont \in \Kont &::= \kmt \mid \kar{\mexpr,\menv,\mkont} \mid \kfn{\mval,\mkont} \\
  \menv \in \Env &= \Var \to \Addr \\
  \mstore \in \Store &= \Addr \to \wp(\Value) \\
  \mvar \in \Var &\text{ an infinite set} \\
  \maddr \in \Addr &\text{ an infinite set}
\end{align*}

@@

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar\mvar, \menv), \mstore, \mkont}$
    &
    $\tpl{\mval,\mstore,\mkont}$ if $\mval \in \mstore(\menv(\mvar))$
    \\
% application
    $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mkont}$
    &
    $\tpl{(\mexpri0, \menv), \mstore, \kar{\mexpri,\menv,\mkont}}$
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kar{\mexpr,\menv,\mkont}}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kfn{\mval, \mkont}}$
    \\
% function call
    $\tpl{\mval,\mstore,\kfn{\sclo{\slam{\mvar}{\mexpr}}{\menv}, \mkont}}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \joinone{\mstore}{\maddr}{\mval}, \mkont}$
  \end{tabular}
  \caption{The CESK machine}
  \label{fig:base-semantics}
\end{figure}

@@

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar\mvar, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{\mval,\mstore,\mkont, \mktab, \mmemo}$ if $\mval \in \mstore(\menv(\mvar))$
    \\
% application
    $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{(\mexpri0, \menv), \mstore, \kar{\mexpri1,\menv,\mkont}, \mktab, \mmemo}$
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kar{\mexpr,\menv,\mkont}, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kfn{\mval, \mkont}, \mktab, \mmemo}$
    \\
% function call
    $\tpl{\mval,\mstore,\kfn{\sclo{\slam{\mvar}{\mexpr}}{\menv}, \mkont},\mktab,\mmemo}$
    & % actually do call,
    $\tpl{\mpoint,
          \mstore',
          \krt{\mpoint, \mstore'},
          \joinone{\mktab}{(\mpoint, \mstore')}{\mkont},
          \mmemo}$
\\
    & $\tpl{\mval, \mstore', \mkont, \mktab, \mmemo} \text{ if } \mval \in \mmemo(\mpoint,\mstore')$
    \\ % or, lookup in table
%    & $\tpl{\mval, \mstore', \mkont,\mktab,\mmemo}$ if $\mval \in \mmemo(\mpoint, \mstore')$ \\    
    where & $\mpoint = (\mexpr, \extm{\menv}{\mvar}{\maddr})$ \\
          & $\mstore' = \joinone{\mstore}{\maddr}{\mval}$
    \\
% return
    $\tpl{\mval, \mstore, \krt{\mpoint,\mstore'}, \mktab, \mmemo}$
    &
    $\tpl{\mval, \mstore, \mkont, \mktab, \joinone{\mmemo}{(\mpoint, \mstore')}{\mval}}$
    if $\mkont \in \mktab(\mpoint, \mstore')$
  \end{tabular}
  \caption{The summarizing tabular stack machine}
  \label{fig:summary-semantics}
\end{figure}

@@

We show only the significantly modified rules of the semantics in
figure \ref{fig:frame-semantics}. The other rules simply carry along
the extra $\mframe$ component untouched.
\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar[\mlab]{\mvar}, \menv), \mstore, \mframe, \mkont}$
    &
    $\tpl{\mval,\mstore,\mframe',\mkont}$ if $(\mframe', \mval) \in \lookup(\mstore,\mframe,\menv,\mvar,\mlab)$
    \\
% % application
%     $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mframe, \mkont}$
%     &
%     $\tpl{(\mexpri0, \menv), \mstore, \mframe, \kar{\mexpri1,\menv,\mkont}}$
%     \\
% % argument evaluation
%     $\tpl{\mval, \mstore, \mframe, \kar{\mexpr,\menv,\mkont}}$
%     &
%     $\tpl{(\mexpr, \menv), \mstore, \mframe, \kfn{\mval, \mkont}}$
%     \\
% function call
    $\tpl{\mval,\mstore,\mframe,\kfn{\sclo{\slam{\mvar}{\mexpr}}{\menv}, \mkont}}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \mstore', \mframe', \mkont}$
    \\ where & $(\mstore',\mframe') = \bind(\mstore,\mframe,\maddr,\mvar,\mval)$
  \end{tabular}
  \caption{The CES$\xi$K machine}
  \label{fig:frame-semantics}
\end{figure}

@@

>>>>

\begin{align*}
  \mexpr \in \Expr &= \svar{\mvar} \mid \sapp{\mexpr}{\mexpr} \mid \slam{\mvar}{\mexpr} \\
  \mstate \in \State &= (\Expr \times \Env) \times \Store \times \Kont \\
  \mval \in \Value &::= \vclo{\slam{\mvar}{\mexpr}}{\menv} \\
  \mkont \in \Kont &::= \kmt \mid \kar{\mexpr,\menv,\mkont} \mid \kfn{\mval,\mkont} \\
  \menv \in \Env &= \Var \to \Addr \\
  \mstore \in \Store &= \Addr \to \wp(\Value) \\
  \mvar \in \Var &\text{ an infinite set} \\
  \maddr \in \Addr &\text{ an infinite set}
\end{align*}

@@

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar\mvar, \menv), \mstore, \mkont}$
    &
    $\tpl{\mval,\mstore,\mkont}$ if $\mval \in \mstore(\menv(\mvar))$
    \\
% application
    $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mkont}$
    &
    $\tpl{(\mexpri0, \menv), \mstore, \kar{\mexpri,\menv,\mkont}}$
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kar{\mexpr,\menv,\mkont}}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kfn{\mval, \mkont}}$
    \\
% function call
    $\tpl{\mval,\mstore,\kfn{\vclo{\slam{\mvar}{\mexpr}}{\menv}, \mkont}}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \joinone{\mstore}{\maddr}{\mval}, \mkont}$
  \end{tabular}
  \caption{The CESK machine}
  \label{fig:base-semantics}
\end{figure}

@@

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar\mvar, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{\mval,\mstore,\mkont, \mktab, \mmemo}$ if $\mval \in \mstore(\menv(\mvar))$
    \\
% application
    $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{(\mexpri0, \menv), \mstore, \kar{\mexpri1,\menv,\mkont}, \mktab, \mmemo}$
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kar{\mexpr,\menv,\mkont}, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kfn{\mval, \mkont}, \mktab, \mmemo}$
    \\
% function call
    $\tpl{\mval,\mstore,\kfn{\vclo{\slam{\mvar}{\mexpr}}{\menv}, \mkont},\mktab,\mmemo}$
    & % actually do call,
    $\tpl{\mpoint,
          \mstore',
          \krt{\mpoint, \mstore'},
          \joinone{\mktab}{(\mpoint, \mstore')}{\mkont},
          \mmemo}$
\\
    & $\tpl{\mval, \mstore', \mkont, \mktab, \mmemo} \text{ if } \mval \in \mmemo(\mpoint,\mstore')$
    \\ % or, lookup in table
%    & $\tpl{\mval, \mstore', \mkont,\mktab,\mmemo}$ if $\mval \in \mmemo(\mpoint, \mstore')$ \\    
    where & $\mpoint = (\mexpr, \extm{\menv}{\mvar}{\maddr})$ \\
          & $\mstore' = \joinone{\mstore}{\maddr}{\mval}$
    \\
% return
    $\tpl{\mval, \mstore, \krt{\mpoint,\mstore'}, \mktab, \mmemo}$
    &
    $\tpl{\mval, \mstore, \mkont, \mktab, \joinone{\mmemo}{(\mpoint, \mstore')}{\mval}}$
    if $\mkont \in \mktab(\mpoint, \mstore')$
  \end{tabular}
  \caption{The summarizing tabular stack machine}
  \label{fig:summary-semantics}
\end{figure}

@@

We show only the significantly modified rules of the semantics in
figure \ref{fig:frame-semantics}. The other rules simply carry along
the extra $\mframe$ component untouched.
\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar[\mlab]{\mvar}, \menv), \mstore, \mframe, \mkont}$
    &
    $\tpl{\mval,\mstore,\mframe',\mkont}$ if $(\mframe', \mval) \in \lookup(\mstore,\mframe,\menv,\mvar,\mlab)$
    \\
% % application
%     $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mframe, \mkont}$
%     &
%     $\tpl{(\mexpri0, \menv), \mstore, \mframe, \kar{\mexpri1,\menv,\mkont}}$
%     \\
% % argument evaluation
%     $\tpl{\mval, \mstore, \mframe, \kar{\mexpr,\menv,\mkont}}$
%     &
%     $\tpl{(\mexpr, \menv), \mstore, \mframe, \kfn{\mval, \mkont}}$
%     \\
% function call
    $\tpl{\mval,\mstore,\mframe,\kfn{\vclo{\slam{\mvar}{\mexpr}}{\menv}, \mkont}}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \mstore', \mframe', \mkont}$
    \\ where & $(\mstore',\mframe') = \bind(\mstore,\mframe,\maddr,\mvar,\mval)$
  \end{tabular}
  \caption{The CES$\xi$K machine}
  \label{fig:frame-semantics}
\end{figure}

@@

Turning this into a table-based semantics involves making prompts a
point of indirection, just like function calls. Memoization also gets
a new context to consider, calling a continuation, because composable
continuations act like functions. Figure \ref{fig:shift-reset-table0}
has what one might naturally write using just this
information. Unfortunately, since $\rt$ frames contain a store, and
continuations can now appear in the store, this introduces a
circularity that could cause the analysis to never terminate. For
example, a loop that continually captures continuations would
introduce unboundedly many continuation values due to the ever-growing
store. This means the store is not finite height, and there may be no
fixed point.

\paragraph{Possible fixes to non-termination:} The simplest fix to the
circularity would be to strip stores out of the context of the $\rt$
frame in a captured continuation, and additionally consider contexts
with a stripped store an abstraction of the same context with any
store. This is a fairly brutal approximation of a captured
continuation, so an alternative is to make this tunable with our
precision-tuning friend, $\alloc$. Then, instead of entirely removing
the store component of the $\rt$ context, we can replace it with an
address of possible stores that it would approximate. The first
brute-force approach is an instantiation of this, where the allocated
address is always the same.

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% Reset
    $\tpl{(\sreset{\mexpr}, \menv), \mstore, \mkont, \mmkont, \mktab, \mmemo}$
    &
    $\tpl{\mpoint, \mstore, \kmt, \kprompt{\mctx}, \joinone{\mktab}{\mctx}{(\mkont,\mmkont)}, \mmemo}$
    \\
    where & $\mpoint = (\mexpr, \menv)$, $\mctx = (\mpoint, \mstore)$.
    \\
% Pop prompt
    $\tpl{\mval, \mstore, \kmt, \kprompt{\mctx}, \mktab, \mmemo}$
    &
    $\tpl{\mval, \mstore, {\mkont}, {\mmkont}, \mktab, \joinone{\mmemo}{\mctx}{\mval}}$
    if $(\mkont,\mmkont) \in \mktab(\mctx)$
    \\
% Shift
    $\tpl{(\sshift{\mvar}{\mexpr}, \menv), \mstore, \mkont, \mmkont, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \joinone{\mstore}{\maddr}{\vcomp{\mkont}},\kmt,\mmkont,\mktab,\mmemo}$
    \\
% continuation call
    $\tpl{\mval,\mstore,\kfn{\vcomp{\mkont'}, \mkont}, \mmkont, \mktab, \mmemo}$
    &
    $\tpl{\mval, \mstore, \mkont', \kprompt{\mctx}, \mktab',\mmemo}$
    \\
    where & $\mctx = (\mkont, \mval, \mstore')$ \\
          & $\mktab' = \joinone{\mktab}{\mctx}{(\mkont,\mmkont)}$
    \\
% Memoized continuation call
% continuation call
    $\tpl{\mval,\mstore,\kfn{\vcomp{\mkont'}, \mkont}, \mmkont, \mktab, \mmemo}$
    &
    $\tpl{\mval', \mstore, \mkont, \mmkont, \mktab,\mmemo}$ if $\mval' \in \mmemo(\mkont',\mval,\mstore)$
  \end{tabular}  
  \caption{Faulty table-based semantics for shift/reset}
  \label{fig:shift-reset-table0}
\end{figure}

The new indirection possibility changes $\Context$ to also include
$\maddr$ where there was previously a $\mstore$, and $\KTab$ now
additionally maps $\Addr$ to a set of stores. We introduce three new
metafunctions to facilitate the new approximation:

\newcommand{\replacectx}{\mathit{replace}\mstore}
\newcommand{\addstore}{\mathit{add}\mstore}
\begin{align*}
  \returns(\mktab, ((\mexpr,\menv), \mstore)) &= \mktab((\mexpr,\menv), \mstore) \\
  \returns(\mktab, ((\mexpr,\menv), \maddr)) &=
    \bigcup\setbuild{\mktab((\mexpr,\menv),\mstore)}{\mstore \in \mktab(\maddr)}
  \\[2pt]
  \memoize(\mmemo, \mktab, ((\mexpr,\menv), \mstore), \mval) &=
    \joinone{\mmemo}{((\mexpr,\menv), \mstore)}{\mval} \\
  \memoize(\mmemo, \mktab, ((\mexpr,\menv), \maddr), \mval) &=
    \mmemo \sqcup \bigsqcup\limits_{\mstore \in \mktab(\maddr)}\snglm{((\mexpr,\menv), \mstore)}{\mval}
  \\[2pt]
  \approximate(\mkont, \mktab, \maddr) &= (\replacectx(\mkont), \addstore{\mkont}) \\
  \text{where }
   \replacectx(\krt{((\mexpr,\menv),\_)}) &= \krt{((\mexpr,\menv),\maddr)} \\
   \replacectx(\kar{\mexpr,\menv,\mkont}) &= \kar{\mexpr,\menv,\replacectx(\mkont)} \\
   \replacectx(\kfn{\mval, \mkont}) &= \kfn{\mval,\replacectx(\mkont)}
  \\[2pt]
   \addstore(\krt{((\mexpr,\menv),\mstore)}) &= \joinone{\mktab}{\maddr}{\mstore} \\
   \addstore(\krt{((\mexpr,\menv),\maddr')}) &= \joinm{\mktab}{\maddr}{\mktab(\maddr')} \\
   \addstore(\kar{\mexpr,\menv,\mkont}) &= \addstore(\mkont) \\
   \addstore(\kfn{\mval, \mkont}) &= \addstore(\mkont)
\end{align*}

\begin{figure}
  \centering
% Shift
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
    $\tpl{(\sshift{\mvar}{\mexpr}, \menv), \mstore, \mkont, \mmkont, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \joinone{\mstore}{\maddr}{\vcomp{\mkont'}},\kmt,\mmkont,\mktab',\mmemo}$
    \\
    where & $\maddr = \alloc(\varsigma)$, $(\mkont',\mktab') = \approximate(\mkont, \mktab, \maddr)$
\\
% Return
   $\tpl{\mval, \mstore, \krt{\mctx}, \mmkont, \mktab, \mmemo}$
   &
   $\tpl{\mval, \mstore, \mkont, \mmkont, \mktab, \mmemo'}$
   if $\mkont \in \returns(\mktab, \mctx)$
   \\
   where & $\mmemo' = \memoize(\mmemo, \mktab, \mctx, \mval)$
  \end{tabular}
  \caption{Fixed table-based semantics for shift/reset}
  \label{fig:shift-reset-table1}
\end{figure}
